---
title: "Auditing GFA traversals"
author: "Anestis Gkanogiannis"
date: "20 August 2026"
package: panpathAudit
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 2
        number_sections: true
vignette: >
  %\VignetteIndexEntry{Auditing GFA traversals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-options, include = FALSE}
knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

`panpathAudit` tests whether source FASTA sequences are preserved by named
traversals in GFA files. It reports exact recovery, divergence coordinates,
digests, and optional base-level statistics. Results and diagnostics use
Bioconductor `DataFrame` containers for integration with R workflows.
The package brings sequence-preservation auditing into reproducible R and
Bioconductor analyses while retaining the bounded native implementation used
by the command-line reference tool.

Tools such as `gfatools` and `odgi` provide broader graph manipulation and
analysis. `panpathAudit` instead provides a focused, reproducible audit of
sequence preservation. It does not validate graph topology. Source code and
issues are available from the [package repository][repository].

# Installation

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("panpathAudit")
```

```{r setup}
library(panpathAudit)
```

# Input scope

Inputs may be plain or gzip-compressed FASTA and GFA files. The package reads
GFA `P` paths and `W` walks. Path overlaps must be blunt (`*` or `0M`), and
sequence symbols must use the IUPAC nucleotide alphabet. Graph topology is not
validated.

# Basic audit

```{r basic}
fasta <- system.file("extdata", "example.fa", package = "panpathAudit")
gfa <- system.file("extdata", "example.gfa", package = "panpathAudit")
result <- auditGFA(fasta, gfa)
auditOutcomes(result)[, c("identifier", "status", "record_type")]
```

The audit checks reconstructed sequence, not graph topology. Range columns
explicitly marked as zero-based retain GFA conventions. Other displayed
positions are one-based.

The result separates traversal outcomes from summary and provenance data.

```{r result-structure}
names(auditOutcomes(result))
auditSummary(result)
auditProvenance(result)[c("gfa_version", "path_records", "walk_records")]
```

# Walks and mappings

`W` identifiers use `sample#haplotype#sequence`. FASTA headers can be mapped
with one row per file.

| Mode | Required fields | Identifier |
|:-----|:----------------|:-----------|
| `exact` | none | FASTA record name |
| `pansn` | `sample`, numeric `haplotype` | `sample#haplotype#name` |
| `prefix` | `prefix` | `prefix` plus record name |

```{r walks}
walk_fasta <- system.file("extdata", "walk.fa", package = "panpathAudit")
walk_gfa <- system.file("extdata", "walk.gfa", package = "panpathAudit")
walk_result <- auditGFA(walk_fasta, walk_gfa)
auditOutcomes(walk_result)[, c(
    "identifier", "status", "source_start_0based",
    "source_end_0based_exclusive", "addressed_source_bases"
)]
```

For an unmapped header such as `chr1`, use:

```{r mapping}
mapped_fasta <- tempfile(fileext = ".fa")
writeLines(c(">chr1", "ACGTGT"), mapped_fasta)
mapping <- data.frame(
    mode = "pansn", sample = "sample", haplotype = "1", prefix = ""
)
mapped_result <- auditGFA(mapped_fasta, walk_gfa, mapping = mapping)
auditOutcomes(mapped_result)[, c("identifier", "status", "record_type")]
```

# Comprehensive statistics

```{r comprehensive}
detailed <- auditGFA(fasta, gfa, statistics = "comprehensive")
auditStatistics(detailed)$global
auditTelemetry(detailed)
```

Alignment work is bounded by `alignmentMaxCells` and `memoryMiB`. An unavailable
alignment remains a completed outcome and records its reason in
`alignment_status`.

# Diagnostics

Argument, input, and operational errors inherit from `panpathAudit_error` and
carry a `diagnostics` `DataFrame`.

```{r diagnostics}
diagnostics <- tryCatch(
    auditGFA(file.path(tempdir(), "missing.fa"), gfa),
    panpathAudit_error = function(error) error$diagnostics
)
diagnostics[, c("category", "code", "message")]
```

# References {.unnumbered}

- [Graphical Fragment Assembly format specification][gfa-spec]
- [Pangenome Sequence Naming specification][pansn-spec]
- [gfatools][gfatools]
- [odgi][odgi]

[repository]: https://github.com/gkanogiannis/panpathAudit
[gfa-spec]: https://github.com/GFA-spec/GFA-spec
[pansn-spec]: https://github.com/pangenome/PanSN-spec
[gfatools]: https://github.com/lh3/gfatools
[odgi]: https://github.com/pangenome/odgi

The underlying audit contract is described in Gkanogiannis (2026),
*Panpath-Audit: Sequence-Level Verification of Embedded Pangenome Paths Against
Source Assemblies*,
DOI 10.20944/preprints202608.1872.v1.

# Session information

```{r session-info}
sessionInfo()
```
