---
title: "Auditing GFA traversals"
author: "Anestis Gkanogiannis"
package: panpathAudit
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Auditing GFA traversals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE}
library(panpathAudit)
```

# Input scope

Inputs may be plain or gzip-compressed FASTA and GFA files. The package reads
GFA `P` paths and `W` walks. Path overlaps must be blunt (`*` or `0M`), and
sequence symbols must use the IUPAC nucleotide alphabet. Graph topology is not
validated.

# Basic audit

```{r basic}
fasta <- system.file("extdata", "example.fa", package = "panpathAudit")
gfa <- system.file("extdata", "example.gfa", package = "panpathAudit")
result <- auditGFA(fasta, gfa)
auditOutcomes(result)[, c("identifier", "status", "record_type")]
```

The audit checks reconstructed sequence, not graph topology. Range columns
explicitly marked as zero-based retain GFA conventions. Other displayed
positions are one-based.

The result separates traversal outcomes from summary and provenance data.

```{r result-structure}
names(auditOutcomes(result))
auditSummary(result)
auditProvenance(result)[c("gfa_version", "path_records", "walk_records")]
```

# Walks and mappings

`W` identifiers use `sample#haplotype#sequence`. FASTA headers can be mapped
with one row per file.

| Mode | Required fields | Identifier |
|:-----|:----------------|:-----------|
| `exact` | none | FASTA record name |
| `pansn` | `sample`, numeric `haplotype` | `sample#haplotype#name` |
| `prefix` | `prefix` | `prefix` plus record name |

```{r walks}
walk_fasta <- system.file("extdata", "walk.fa", package = "panpathAudit")
walk_gfa <- system.file("extdata", "walk.gfa", package = "panpathAudit")
walk_result <- auditGFA(walk_fasta, walk_gfa)
auditOutcomes(walk_result)[, c(
    "identifier", "status", "source_start_0based",
    "source_end_0based_exclusive", "addressed_source_bases"
)]
```

For an unmapped header such as `chr1`, use:

```{r mapping, eval=FALSE}
mapping <- data.frame(
    mode = "pansn", sample = "sample", haplotype = "1", prefix = ""
)
auditGFA("assembly.fa", "graph.gfa", mapping = mapping)
```

# Comprehensive statistics

```{r comprehensive}
detailed <- auditGFA(fasta, gfa, statistics = "comprehensive")
auditStatistics(detailed)$global
auditTelemetry(detailed)
```

Alignment work is bounded by `alignmentMaxCells` and `memoryMiB`. An unavailable
alignment remains a completed outcome and records its reason in
`alignment_status`.

# Diagnostics

Argument, input, and operational errors inherit from `panpathAudit_error` and
carry a `diagnostics` `DataFrame`.

```{r diagnostics, eval=FALSE}
tryCatch(
    auditGFA("missing.fa", "graph.gfa"),
    panpathAudit_error = function(error) error$diagnostics
)
```

# Session information

```{r session-info}
sessionInfo()
```
