---
title: "Auditing GFA traversals"
author: "Anestis Gkanogiannis"
date: "`r format(Sys.Date(), '%d %B %Y')`"
package: panpathAudit
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 2
        number_sections: true
vignette: >
  %\VignetteIndexEntry{Auditing GFA traversals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-options, include = FALSE}
knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    collapse = TRUE,
    comment = "#>",
    # BiocStyle's crop hook needs magick; FALSE does not disable it, only NULL
    crop = NULL
)

cap1 <- paste(
    "Anatomy of a reverse-strand divergence. The highlighted column and",
    "both position rulers are taken from the audit result above; the",
    "segment blocks are the graph defined in this section.")
cap2 <- paste(
    "Walk coverage and the two coordinate conventions. Edge numbers are",
    "zero-based and half-open; base numbers are one-based. Ranges and",
    "totals are taken from the outcome row above.")
cap3 <- paste(
    "How the comprehensive base buckets partition each side. Both bars",
    "are drawn from the values printed above and sum to the reported",
    "source and graph base totals.")
```

# Introduction

A pangenome graph usually claims to embed the assemblies it was built from: each
named `P` path or `W` walk should spell out, base for base, the contig it is
labelled with. That claim is easy to break and easy to miss. Graph construction
and normalization rewrite segments, chunked builds are stitched back together,
coordinates are lifted between references, and files are round-tripped through
several tools before anyone looks at them. The graph stays well formed
throughout, so structural validators stay quiet.

`panpathAudit` answers a narrower question: **is the source sequence still
recoverable from the traversal, and if not, exactly where does it stop
matching?** It reconstructs the sequence spelled by every named traversal,
compares it against the source FASTA, and reports the first divergent base with
its source, traversal and segment coordinates, along with digests and optional
base-level alignment statistics.

Tools such as `gfatools` and `odgi` provide broader graph manipulation and
analysis. `panpathAudit` instead provides a focused, reproducible audit of
sequence preservation. It does not validate graph topology: links, overlaps
beyond blunt joins, and connectivity are outside its scope. Results and
diagnostics use Bioconductor `DataFrame` containers for integration with R
workflows, while the bounded native implementation used by the command-line
reference tool is retained. Source code and issues are available from the
[package repository][repository].

# Installation

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("panpathAudit")
```

You can install the development version of `panpathAudit`:
```r
devtools::install_github("gkanogiannis/panpathAudit")
```
Building the package requires a C++20 compiler and zlib.

```{r setup}
library(panpathAudit)
```

# Input scope

Inputs may be plain or gzip-compressed FASTA and GFA files; compression is
detected from file content, not from the file name. The package reads GFA `P`
paths and `W` walks. Path overlaps must be blunt (`*` or `0M`), and sequence
symbols must use the IUPAC nucleotide alphabet. Graph topology is not validated.

# Basic audit

```{r basic}
fasta <- system.file("extdata", "example.fa", package = "panpathAudit")
gfa <- system.file("extdata", "example.gfa", package = "panpathAudit")
result <- auditGFA(fasta, gfa)
result
```

Printing the object gives the headline counts. The per-traversal detail lives in
the outcomes table.

```{r basic-outcomes}
auditOutcomes(result)[, c("identifier", "status", "record_type")]
```

The result separates traversal outcomes from summary, provenance and telemetry
data. These accessors return `SimpleList` objects, so wrap them in `as.list()`
when you want their contents printed.

```{r result-structure}
as.list(auditSummary(result))
provenance <- auditProvenance(result)
as.list(provenance[c("gfa_version", "path_records", "walk_records")])
```

Provenance also records the resolved input paths, the mapping and resource
options in force, and a `topology_validated` flag that is always `FALSE` — a
reminder of what the audit does not cover.

# Audit statuses

Every source record and every named traversal is accounted for by exactly one of
four statuses. All four are completed outcomes, not errors.

| Status | Meaning |
|:-------|:--------|
| `IDENTICAL` | The traversal spells the source sequence exactly. |
| `DIVERGENT` | Both sides exist but differ; see divergence columns. |
| `MISSING_PATH` | A source record has no traversal of that name. |
| `MISSING_SOURCE` | A traversal has no source record of that name. |

A single audit can report all four at once. The following graph embeds one
contig faithfully, one with a substitution, carries a path with no source, and
leaves one source record unrepresented.

```{r mixed-setup}
workdir <- tempfile("panpathAudit-demo-")
dir.create(workdir)

mixed_fasta <- file.path(workdir, "mixed.fa")
mixed_gfa <- file.path(workdir, "mixed.gfa")

writeLines(c(
    ">same", "ACGT",
    ">changed", "AAAA",
    ">absent", "NN"
), mixed_fasta)

writeLines(c(
    "H\tVN:Z:1.1",
    "S\t1\tACGT",
    "S\t2\tAATA",
    "S\t3\tGG",
    "P\tsame\t1+\t*",
    "P\tchanged\t2+\t*",
    "P\textra\t3+\t*"
), mixed_gfa)
```

```{r mixed-audit}
mixed <- auditGFA(mixed_fasta, mixed_gfa)
auditOutcomes(mixed)[, c(
    "identifier", "status", "record_type", "source_length", "graph_length"
)]
auditSummary(mixed)$counts
```

`source_length` is `NA` for `MISSING_SOURCE` and `graph_length` is `NA` for
`MISSING_PATH`; columns that do not apply to a row are always `NA`. See
`?PanPathAuditResult` for the full column catalogue.

# Reading a divergence

The point of the audit is the `DIVERGENT` row. Consider a source contig `ACGA`
against a two-segment path that spells `AC` + `GT`:

```{r divergent-setup}
divergent_fasta <- file.path(workdir, "divergent.fa")
divergent_gfa <- file.path(workdir, "divergent.gfa")

writeLines(c(">sample", "ACGA"), divergent_fasta)
writeLines(c(
    "H\tVN:Z:1.1",
    "S\t1\tAC",
    "S\t2\tGT",
    "P\tsample\t1+,2+\t*"
), divergent_gfa)
```

```{r divergent-audit}
divergent <- auditGFA(divergent_fasta, divergent_gfa)
auditOutcomes(divergent)[, c(
    "status", "divergence_kind", "source_position_1based", "segment",
    "orientation", "traversal_position_1based", "segment_position_1based"
)]
```

The source reads `ACGA` and the path spells `ACGT`, so the first three bases
agree and the fourth does not. The report locates that base three ways, all of
them **one-based**, as the `_1based` suffixes indicate: it is base 4 of the
source contig, it falls in segment `2` (`GT`), and within that segment it is
base 2 — the `T` the graph supplies where the source has `A`.
`traversal_position_1based` locates that same divergence along the segment *as
the path reads it*. This step is `+`, so the path reads segment `2` forwards and
the two coordinates coincide. On a reverse step they do not, which is why both
are reported; *Reverse traversals* below shows that case.

`divergence_kind` distinguishes three failure shapes:

- `BASE` — both sides supply a base at that position and they differ.
- `END` — one side ran out first.
- `RANGE_LENGTH` — a walk reconstructed to a different length than its
    declared range.

When the graph is exhausted before the source is, no segment can be named:

```{r divergent-end}
truncated_fasta <- file.path(workdir, "truncated.fa")
truncated_gfa <- file.path(workdir, "truncated.gfa")

writeLines(c(">sample", "ACGT"), truncated_fasta)
writeLines(c("H\tVN:Z:1.1", "S\t1\tACG", "P\tsample\t1+\t*"), truncated_gfa)

auditOutcomes(auditGFA(truncated_fasta, truncated_gfa))[, c(
    "divergence_kind", "source_position_1based", "source_length",
    "graph_length", "segment"
)]
```

## Reverse traversals

When a path traverses a segment on the reverse strand, the two position columns
come apart, which is why both are reported. Here the divergence is at source
base 5, inside the reverse-complemented segment `2` (`TTGA`, four bases long):

```{r divergent-reverse}
reverse_fasta <- file.path(workdir, "reverse.fa")
reverse_gfa <- file.path(workdir, "reverse.gfa")

writeLines(c(">sampleB#1#chr1", "ACGTGCAACC"), reverse_fasta)
writeLines(c(
    "H\tVN:Z:1.0",
    "S\t1\tACGT", "S\t2\tTTGA", "S\t3\tCC",
    "P\tsampleB#1#chr1\t1+,2-,3+\t*"
), reverse_gfa)

reverse <- auditGFA(reverse_fasta, reverse_gfa)
auditOutcomes(reverse)[, c(
    "source_position_1based", "segment", "orientation",
    "traversal_position_1based", "segment_position_1based"
)]
```

`traversal_position_1based` is 1 — the divergence is at the first base the path
reads out of that segment — while `segment_position_1based` is 4, because
reading `TTGA` backwards starts at its last base. Use the traversal coordinate
to reason about the path, and the segment coordinate to find the base in the `S`
record.

```{r fig-div, echo=FALSE, fig.dim=c(7.5, 3.8), fig.cap=cap1}
plot_divergence <- function(row) {
    src <- "ACGTGCAACC"
    gph <- "ACGTTCAACC"
    segs <- list(
        list(name = "1", orient = "+", from = 1, to = 4, seq = "ACGT"),
        list(name = "2", orient = "-", from = 5, to = 8, seq = "TTGA"),
        list(name = "3", orient = "+", from = 9, to = 10, seq = "CC"))
    pos <- row$source_position_1based
    n <- nchar(src)
    pal <- palette.colors(8, "Okabe-Ito")
    hit <- pal[7]; blue <- pal[3]; grey <- "#E9E9E9"; ink <- "#222222"

    old <- par(mar = c(0.2, 0.2, 0.2, 0.2))
    on.exit(par(old))
    plot.new()
    plot.window(xlim = c(-2.9, n + 0.3), ylim = c(0, 7.1))

    cell <- function(i, y0, y1, ch, fill, col) {
        rect(i - 1, y0, i, y1, col = fill, border = "white", lwd = 2)
        text(i - 0.5, (y0 + y1) / 2, ch, col = col, font = 2, cex = 1.05)
    }
    lab <- function(y, txt, cex = 0.9, col = ink)
        text(-0.25, y, txt, adj = c(1, 0.5), cex = cex, col = col)

    for (i in seq_len(n)) {
        on <- i == pos
        cell(i, 6.0, 6.9, substr(src, i, i), if (on) hit else grey,
            if (on) "white" else ink)
        cell(i, 4.9, 5.8, substr(gph, i, i), if (on) hit else grey,
            if (on) "white" else ink)
    }
    lab(6.45, "source")
    lab(5.35, "graph")

    for (sg in segs) {
        rev <- sg$orient == "-"
        rect(sg$from - 1, 3.9, sg$to, 4.6,
            col = if (rev) blue else "#D9D9D9", border = "white", lwd = 2)
        text((sg$from - 1 + sg$to) / 2, 4.25,
            sprintf("%s (%s)  %s", sg$name, sg$orient, sg$seq),
            col = if (rev) "white" else ink, cex = 0.85, font = 2)
    }
    lab(4.25, "segments")

    rs <- segs[[2]]
    k <- rs$to - rs$from + 1
    for (j in seq_len(k)) {
        x <- rs$from - 1 + j - 0.5
        on <- (rs$from + j - 1) == pos
        text(x, 3.25, j, cex = 0.9, col = if (on) hit else ink,
            font = if (on) 2 else 1)
        text(x, 2.65, k - j + 1, cex = 0.9, col = if (on) hit else ink,
            font = if (on) 2 else 1)
    }
    lab(3.25, "traversal position", 0.8)
    lab(2.65, "segment position", 0.8)
    arrows(rs$to - 0.1, 2.1, rs$from - 0.9, 2.1, length = 0.08,
        col = blue, lwd = 2)
    text((rs$from - 1 + rs$to) / 2, 1.75,
        sprintf("%s reverse-complemented = TCAA", rs$seq),
        cex = 0.8, col = blue)

    segments(pos - 0.5, 4.9, pos - 0.5, 4.62, col = hit, lwd = 2)
    text(n / 2, 0.85, sprintf(
        "divergence: source %d  |  segment %s  |  orientation %s",
        pos, row$segment, row$orientation), cex = 0.88, col = ink)
    text(n / 2, 0.3, sprintf(
        "traversal position %d  =  segment position %d",
        row$traversal_position_1based, row$segment_position_1based),
        cex = 0.88, col = hit, font = 2)
}

plot_divergence(as.data.frame(auditOutcomes(reverse)))
```

## Digests

Every row also carries `source_sha256`, `graph_sha256` and their BLAKE3
counterparts (plus `addressed_*` variants for walks). They let you confirm an
`IDENTICAL` verdict independently, or compare an audit against another tool's
output, without re-reading either file.

```{r digests}
as.data.frame(auditOutcomes(divergent)[, c("source_sha256", "graph_sha256")])
```

# Walks and mappings

`W` records identify a sequence as `sample#haplotype#sequence` and address a
half-open, zero-based range of it. Assembly FASTA headers are rarely written in
that form, so without help the two sides do not meet: the same file audited
against a walk graph reports the source as `MISSING_PATH` and the walk as
`MISSING_SOURCE`.

```{r mapping-unmapped}
walk_gfa <- system.file("extdata", "walk.gfa", package = "panpathAudit")

plain_fasta <- file.path(workdir, "chr1.fa")
writeLines(c(">chr1", "ACGTGT"), plain_fasta)

auditOutcomes(auditGFA(plain_fasta, walk_gfa))[, c(
    "identifier", "status", "record_type"
)]
```

A mapping supplies the missing naming convention, with one row per FASTA file.

| Mode | Required fields | Identifier |
|:-----|:----------------|:-----------|
| `exact` | none | FASTA record name |
| `pansn` | `sample`, numeric `haplotype` | `sample#haplotype#name` |
| `prefix` | `prefix` | `prefix` plus record name |

```{r mapping}
mapping <- data.frame(
    mode = "pansn", sample = "sample", haplotype = "1", prefix = ""
)
mapped <- auditGFA(plain_fasta, walk_gfa, mapping = mapping)
auditOutcomes(mapped)[, c("identifier", "status", "record_type")]
```

Because a walk addresses only part of its source, a walk outcome reports both
the range it covers and how much of the contig that range actually reaches.

```{r walks}
walk_fasta <- system.file("extdata", "walk.fa", package = "panpathAudit")
walk_result <- auditGFA(walk_fasta, walk_gfa)
auditOutcomes(walk_result)[, c(
    "identifier", "status", "source_start_0based",
    "source_end_0based_exclusive", "addressed_source_bases", "source_length"
)]
```

Mind the two conventions in that one row. `source_start_0based` and
`source_end_0based_exclusive` retain GFA's **zero-based, half-open** ranges, so
`0` and `6` span the whole six-base contig. The graph reaches only four of those
bases: this example holds two walks, covering `[0, 2)` and `[4, 6)`, leaving the
two bases in between unaddressed. Every other displayed position, including all
the divergence coordinates above, is one-based.

```{r fig-cov, echo=FALSE, fig.dim=c(6.8, 3.6), fig.cap=cap2}
plot_coverage <- function(
    row, seq = "ACGTGT", walks = list(c(0, 2), c(4, 6))) {
    n <- nchar(seq)
    pal <- palette.colors(8, "Okabe-Ito")
    green <- pal[4]; grey <- "#E9E9E9"; ink <- "#222222"; mute <- "#777777"
    covered <- rep(FALSE, n)
    for (w in walks) covered[(w[1] + 1):w[2]] <- TRUE

    old <- par(mar = c(0.2, 0.2, 0.2, 0.2))
    on.exit(par(old))
    plot.new()
    plot.window(xlim = c(-3.0, n + 0.4), ylim = c(0, 6.2))

    for (i in seq_len(n)) {
        on <- covered[i]
        rect(i - 1, 3.5, i, 4.5, col = if (on) green else grey,
            border = "white", lwd = 2)
        text(i - 0.5, 4.0, substr(seq, i, i), font = 2, cex = 1.05,
            col = if (on) "white" else mute)
    }
    text(-0.3, 4.0, "contig", adj = c(1, 0.5), cex = 0.85, col = ink)

    for (e in 0:n) {
        segments(e, 4.55, e, 4.85, col = mute, lwd = 1)
        text(e, 5.15, e, cex = 0.85, col = ink)
    }
    text(-0.3, 5.15, "0-based edges", adj = c(1, 0.5), cex = 0.85, col = ink)
    text(n / 2, 5.75, "zero-based, half-open: boundaries between bases",
        cex = 0.82, col = mute)

    for (i in seq_len(n)) text(i - 0.5, 3.1, i, cex = 0.85, col = ink)
    text(-0.3, 3.1, "1-based bases", adj = c(1, 0.5), cex = 0.85, col = ink)

    for (w in walks) {
        segments(w[1] + 0.06, 2.45, w[2] - 0.06, 2.45, col = green, lwd = 3)
        text(mean(w), 2.05, sprintf("W [%d, %d)", w[1], w[2]),
            cex = 0.8, col = green)
    }
    gap <- which(!covered)
    segments(min(gap) - 0.94, 2.45, max(gap) - 0.06, 2.45,
            col = mute, lwd = 2, lty = 3)
    text(mean(gap) - 0.5, 2.05, sprintf("not_embedded (%d)", length(gap)),
        cex = 0.8, col = mute)
    text(-0.3, 2.45, "walk ranges", adj = c(1, 0.5), cex = 0.85, col = mute)

    segments(row$source_start_0based + 0.06, 1.3,
            row$source_end_0based_exclusive - 0.06, 1.3,
            col = ink, lwd = 1.5)
    text(n / 2, 0.9, sprintf("outcome row spans [%d, %d)",
        row$source_start_0based, row$source_end_0based_exclusive),
        cex = 0.8, col = ink)
    text(n / 2, 0.3, sprintf(
        "addressed_source_bases = %d   of   source_length = %d",
        row$addressed_source_bases, row$source_length),
        cex = 0.88, col = ink, font = 2)
}

plot_coverage(as.data.frame(auditOutcomes(walk_result)))
```

# Comprehensive statistics

`statistics = "comprehensive"` adds a base-level alignment of every divergent
pair and aggregates the result globally, by source and by record type.

```{r comprehensive}
detailed <- auditGFA(mixed_fasta, mixed_gfa, statistics = "comprehensive")
names(auditStatistics(detailed))

global <- auditStatistics(detailed)$global
global$bases
```

The buckets account for every base the audit saw, on both sides. `matched` and
`substituted` count aligned *columns*, each pairing one source base with one
graph base; the remaining buckets count bases on one side only.

- `matched`, `substituted` — aligned columns that agree or disagree.
- `source_only`, `graph_only` — insertions and deletions in the alignment.
- `not_embedded` — source bases no walk range addresses.
- `missing_path`, `missing_source` — bases in records with no counterpart.
- `unaligned_divergent_*` — divergent bases left unaligned by a resource limit.

Reading the example: of the 10 source bases, 7 align and match, 1 aligns and is
substituted, and 2 sit in `absent`, which has no path. Of the 10 graph bases,
the same 8 align and 2 sit in `extra`, which has no source. Per-row alignment
detail sits alongside in the outcomes table.

```{r fig-part, echo=FALSE, fig.dim=c(6.8, 2.9), fig.cap=cap3}
plot_partition <- function(bases, global) {
    pal <- palette.colors(8, "Okabe-Ito")
    spec <- list(
        matched        = list(col = pal[4], lab = "matched"),
        substituted    = list(col = pal[7], lab = "substituted"),
        source_only    = list(col = pal[2], lab = "source_only"),
        graph_only     = list(col = pal[2], lab = "graph_only"),
        not_embedded   = list(col = "#BBBBBB", lab = "not_embedded"),
        missing_path   = list(col = pal[3], lab = "missing_path"),
        missing_source = list(col = pal[8], lab = "missing_source"))
    sides <- list(
        source = c("matched", "substituted", "source_only",
                "not_embedded", "missing_path"),
        graph  = c("matched", "substituted", "graph_only", "missing_source"))
    totals <- c(source = global$source_bases, graph = global$graph_bases)
    used <- character()

    old <- par(mar = c(0.2, 0.2, 0.2, 0.2))
    on.exit(par(old))
    plot.new()
    xmax <- max(totals)
    plot.window(xlim = c(-2.7, xmax + 0.3), ylim = c(-1.7, 3.0))
    ys <- c(source = 2.05, graph = 0.85)

    for (nm in names(sides)) {
        x <- 0
        y <- ys[[nm]]
        for (k in sides[[nm]]) {
            v <- bases[[k]]
            if (is.null(v) || v <= 0) next
            used <- c(used, k)
            rect(x, y - 0.38, x + v, y + 0.38, col = spec[[k]]$col,
                border = "white", lwd = 2)
            text(x + v / 2, y, v, col = "white", font = 2, cex = 0.95)
            x <- x + v
        }
        text(-0.3, y, sprintf("%s bases", nm), adj = c(1, 0.5), cex = 0.9)
        text(xmax + 0.25, y, totals[[nm]], adj = c(0, 0.5), cex = 0.85,
            col = "#777777")
    }

    used <- unique(used)
    per <- xmax / max(length(used), 1)
    for (i in seq_along(used)) {
        k <- used[[i]]
        x0 <- (i - 1) * per
        rect(x0, -0.75, x0 + 0.42, -0.42, col = spec[[k]]$col,
            border = "white")
        text(x0 + 0.58, -0.58, spec[[k]]$lab, adj = c(0, 0.5), cex = 0.72)
    }
    text(xmax / 2, -1.35,
        "matched and substituted are aligned columns, on both sides",
        cex = 0.75, col = "#777777")
}

plot_partition(global$bases, global)
```

```{r comprehensive-rows}
auditOutcomes(detailed)[, c(
    "identifier", "status", "matched", "substituted", "edit_distance",
    "alignment_status"
)]
```

`alignment_status` records why a row does or does not carry alignment numbers:
`not_required` for an identical pair, `complete` for a finished alignment,
`not_applicable` where one side is absent, and `limit_exceeded` or
`memory_limit_exceeded` where a bound stopped the work. Alignment is bounded by
`alignmentMaxCells` and `memoryMiB`; an unavailable alignment remains a
completed outcome and is reported in the `unaligned_divergent_*` buckets rather
than failing the audit. `auditTelemetry()` reports what the run actually used.

```{r telemetry}
as.list(auditTelemetry(detailed))
```

Set `threads` to audit records in parallel; results are deterministic and do not
depend on the thread count.

# Diagnostics

Argument, input, and operational errors raise `panpathAudit_argument_error`,
`panpathAudit_input_error`, and `panpathAudit_operational_error` respectively.
All three inherit from `panpathAudit_error`, so one handler catches any of them,
and each carries a `diagnostics` `DataFrame` rather than only a message.

```{r diagnostics}
diagnostics <- tryCatch(
    auditGFA(file.path(tempdir(), "missing.fa"), gfa),
    panpathAudit_error = function(error) error$diagnostics
)
diagnostics[, c("category", "code", "message")]
```

Alongside `category`, `code` and `message`, the table carries `path`, `line`,
`identifier`, `position` and `symbol` wherever the failure can be pinned to a
location; unrelated fields are `NA`. Diagnostic positions are one-based. A
malformed input reports every problem it found, not just the first, so one call
is enough to see the full picture.

# Citation

```{r citation}
citation("panpathAudit")
```

# References {.unnumbered}

- Gkanogiannis, A. (2026). *Panpath-Audit: Sequence-Level Verification of
    Embedded Pangenome Paths Against Source Assemblies*. Preprints.
    DOI [10.20944/preprints202608.1872.v1][preprint].
- [Graphical Fragment Assembly format specification][gfa-spec]
- [Pangenome Sequence Naming specification][pansn-spec]
- [gfatools][gfatools]
- [odgi][odgi]

[repository]: https://github.com/gkanogiannis/panpathAudit
[preprint]: https://doi.org/10.20944/preprints202608.1872.v1
[gfa-spec]: https://github.com/GFA-spec/GFA-spec
[pansn-spec]: https://github.com/pangenome/PanSN-spec
[gfatools]: https://github.com/lh3/gfatools
[odgi]: https://github.com/pangenome/odgi

# Session information

```{r session-info}
sessionInfo()
```
